GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-11 manuscript.tex
text improvements through to Conclusion

The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. Currently, all of the elements of GeoSPARQL 1.0, including the profile declaration, 
can be found within the SWG's working online repository \footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.
The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 1.1 releases' updated
resources, currently expected in mid-2021.

\subsection{New geometry literals}\label{sec:newliterals}
GeoSPARQL 1.1 introduces three new litergeometry serializations:

\begin{enumerate}
    \item GeoJSON (Geo- JavaScript Object Notation)\cite{butler2016geojson}
    \item KML (Keyhole Markup Language)\cite{nolan2014keyhole} 
    \item DGGS (Discrete Global Grid System)\cite{sahr1998discrete}
\end{enumerate}

The first two of these are ``obvious'' in the sence that that they have been expected/requested by the SDWWG and many users
of GeoSPARQL for a long time, due to those format's popularity. an example of a GeoJSON geometry serialization is given below.

\small
\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily]
"""{"type":"Point", "coordinates":[-83.38,33.95]}"""
^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
\end{lstlisting}
\normalsize
Since the GeoJSON and KML formats are restricted to be represented in the WGS84 SRS only, these restrictions 
also apply to GeoJSON and KML literals. The DGGS WKT literal

@@

\subsection{New spatial functions}\label{sec:newfunctions}
GeoSPARQL 1.1 includes new spatial aggregate functions, which allow for the quick aggregation of different geometry types. 
While spatial aggregate functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregate functions had not yet been introduced into the SPARQL standard 
since SPARQL 1.1 \cite{w3c_sparql_working_group_sparql_2013} was released about one year later. Spatial aggregate functions 
similar to traditional aggregate functions such as AVG,MAX, or MIN allow to aggregate results of geometry queries, e.g., to 
create the union out of a set of given geometry literal results. While calculating these aggregates may also be possible 
outside of the semantic database, the inclusion of the functions provides distinct advantages:

\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (e.g., a single union geometry vs. a set of geometries for which to create union externally)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}

GeoSPARQL 1.1 defines the aggregate functions \emph{geosf:boundingCircle}, calculating a bounding circle around a set of 
geometry, \emph{geosf:centroid} calculating the centroid of the set of geometries, \emph{geosf:concatLines} concatenating 
a set of linestrings that overlap, \emph{geosf:concaveHull} calculating the concave hull of a set of geometries and\emph{geosf:union}, 
calculating the union of a set of geometreis, in addition to the \emph{geosf:envelope} and \emph{geosf:convexHull} functions 
already defined in the GeoSPARQL 1.0 standard. Since SPARQL 1.1 these fucntions were also usable as aggregate functions.

Spatial aggregate function definitions are accompanied by the functions \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}}, 
\href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}}, \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}} 
and \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}}, \href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}}, 
\href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} which allow to retrieve the minimum and maximum 
coordinates of a geometry respectively.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{geo:SpatialMeasure}. 
This class represents a measurement such as a volume, length, or area associated with a measurement amount and a measurement 
unit. It acts as a range of three newly defined properties \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasArea}, 
\href{http://www.opengis.net/ont/geosparql#hasLength}{geo:hasLength}, \href{http://www.opengis.net/ont/geosparql#hasVolume}{geo:hasVolume} 
which make these attributes of a geometry better accessible using SPARQL. Finally, GeoSPARQL 1.1 adds the 
\href{http://www.opengis.net/ont/geosparql#inSRS}{geo:inSRS} property, which allows querying the SRS of a given geometry 
without the need to analyze geometry literal serializations. The property allows for the definition of a SRS as a URI. It 
paves the way for a future definition of SRS fully in RDF as anticipated for GeoSPARQL 2.0.

@@

\section{Modernizing the documentation of GeoSPARQL}\label{sec:documentation}

\subsection{Profile}\label{sec:profile}

>>>>

The profile declaration for GeosPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeosPARQL 1.1 will be published at the same time as all, or most, of the 
1.1 releases' updated resources, currently expected in mid-2021. The 1.1 releases' resources are:

\begin{enumerate}
    \item a profile declaration
    \item a specification document
    \item an RDF/OWL ontology document
    \item a Functions \& Rules vocabulary, derived from the ontology
    \item a Simple Features feature types vocabulary
    \item a set of RIF\cite{kifer2013rif} rules
    \item SHACL\cite{knublauch_shapes_2017} shapes for RDF data validation 
\end{enumerate}

Both the 1.0 and the 1.1 resources are currently avaialbe from the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}.

\subsection{New geometry literals}\label{sec:newliterals}
Three new geometry serializations are introduced:

\begin{enumerate}
    \item \textbf{GeoJSON} (Geo- JavaScript Object Notation)\cite{butler2016geojson}
    \item \textbf{KML} (Keyhole Markup Language)\cite{nolan2014keyhole} 
    \item \textbf{DGGS} (Discrete Global Grid System)\cite{sahr1998discrete}
\end{enumerate}

An example of a point's GeoJSON geometry serialization is given below, followed by an unrelated simple polygon \textit{AusPIX}
\footnote{\url{https://w3id.org/dggs/auspix}} DGGS geometry serialization.

\small
\begin{lstlisting}[caption=GeoJSON geometry serialization example,label=lst:geojsonliteral,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
"""{"type":"Point", "coordinates":[-83.38,33.95]}"""
^^<http://www.opengis.net/ont/geosparql#geoJSONLiteral>
\end{lstlisting}

\begin{lstlisting}[caption=AusPIX DGGS geometry serialization example,label=lst:geodggsWktliteral,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
"""<https://w3id.org/dggs/auspix> DirectedOrdinateList 
(R3231 R3234 R3235 R3238 R3243 R3246)"""
^^<http://www.opengis.net/ont/geosparql#dggsWktLiteral>
\end{lstlisting}
\normalsize

The first two of these new serializations have been much anticipated and were requested by the SDWWG and many users of 
GeoSPARQL, due to those formats' popularity.

@@

\subsection{New spatial functions}\label{sec:newfunctions}
While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
similar to traditional (relation database) aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
queries, for example to create the union of a set of selected serialized geometries. While calculating these aggregates is 
certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:

\begin{enumerate}
    \item No client-side library is needed to create an aggregated geometry result
    \item Fewer and more appropriate results can be returned using a spatial aggregate function (for example, a union geometry result)
    \item Spatial aggregates from different SPARQL endpoints can easily be calculated using federated queries 
\end{enumerate}

In addition to the \emph{geosf:union}, \emph{geosf:envelope} and \emph{geosf:convexHull} functions already defined in GeoSPARQL 1.0, 
for use within SPARQL \texttt{FILTER} operations, GeoSPARQL 1.1 defines \emph{geosf:Union}, \emph{geosf:Envelope} and \emph{geosf:ConvexHull}
as well as \emph{geosf:BoundingCircle}, \emph{geosf:Centroid}, \emph{geosf:ConcatLines} - concatenating a set of overlapping linestrings 
that overlap - and \emph{geosf:ConcaveHull} that can be used to return aggregated results. Listing~\ref{lst:agfunc} gives an example 
of one of these new functions in use.

\small
\begin{lstlisting}[caption=Aggregation Function example SPARQL query,label=lst:agfunc,language=sql,frame=single,basicstyle=\ttfamily,showstringspaces=false]
# returns the circle geometry bounding all the geometries 
# of Feature <x>
SELECT (geosaf:BoundingCircle(?geo) AS ?circ)
WHERE {
    <x> geo:hasGeometry/geo:asWKT ?geo .
}
\end{lstlisting}
\normalsize

New functions to retieve the minimum and maximum values of geometries' coordinates are also added: \emph{geof:minX} \& \emph{geof:maxX},
\emph{geof:minY} \& \emph{geof:maxY} and \emph{geof:minZ} \& \emph{geof:maxZ}.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \texttt{geo:SpatialMeasure}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\texttt{geo:hasArea}, \texttt{geo:hasLength} and \texttt{geo:hasVolume} which make these attributes of a geometry better accessible using 
SPARQL.

These additions address a number of change requests from the SDWWG and the SWG but, given the introduction of a dedicated \textit{measure} class, 
the do more than just provide for a few scalar spatial measures and open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.

The 1.1 release also adds the property \texttt{geo:inSRS}, which allows querying for the SRS of a given geometry without the need to analyze 
geometry serializations. The property allows for the definition of a SRS as an Internationalized Resource Identifier (IRI). It paves the way 
for a future definition of SRS fully in RDF, as anticipated for GeoSPARQL 2.0. Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
of release 1.1.

@@
